---
title: "dplyr introduction"
author: "Max Kruisbrink"
date: "`r Sys.Date()`"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# define knitr paths
base.dir <- "~/Developer/mkruisbrink.github.io/"
base.url <- "/"
fig.path <- "img/rmd/"

# set knitr parameters
opts_knit$set(base.dir = base.dir, base.url = base.url)
opts_chunk$set(fig.path = fig.path) 
```

## The tidyverse

I wrote [a simple introductory to the tidyverse]({% post_url 2022-08-11-tidyverse-introduction %}) of which `dplyr` is a core part.
The tidyverse is a package in R that is - in itself - a collection of individual R packages that all work closely together in the same way (they speak thesame kind of language, if you will). Data import & export, manipulation, exploration and visualization becomes simpler and more elegant by using new and additional syntax which is more intuitive to and readable for us humans.

### Set-up

You can install the tidyverse via **`install.packages("tidyverse")`** and you only have to install it once. Lets load the tidyverse and the data we're going to use.

```{r install and load, warning=FALSE}

# installs the package if required
#install.packages("tidyverse")

# loads the tidyverse
library(tidyverse)

```

### Example data

```{r loading data, warning=FALSE}

# using the starwars dataset
data <- starwars

# for the sake of this introduction we will drop all missing values for now
data <- drop_na(data)

```

## dplyr

>
The pipe
dplyr provides the pipe (`%>%`) operator from magrittr. The `%>%` takes what is on the left hand side and pipes it into the right side as the first argument. You can use the pipe to rewrite multiple operations that you can read left-to-right, top-to-bottom (reading the pipe operator as __“and then”__).

With dplyr you get five key functions that will help you solve most - if not all - of your data manipulation challenges. 

* With `filter()` you can pick observations (rows) based on values.

```{r} 
# normal syntax
filter(data, #data goes here
       sex == "male" & homeworld == "Tatooine") 

# tidy syntax with pipe operator
data %>% # take the data... and then...
  filter(., # the . is where the data gets 'piped' in, the first argument. You can omit this part!
         sex == "male" & homeworld == "Tatooine") 
```

* With `arrange()` you are able to sort rows. If the variable is not numeric, it defaults to alphabetical sorting. The default is ascending order.

```{r}
data %>% #and then...
  filter(sex == "male" & homeworld == "Tatooine") %>% #and then...
  arrange(desc(height)) #sort descending
```


* With `select()` you can pick variables (columns) by their names. You can specify the variable names or provide the numbers, such as variable 1 through 4: `1:4`. 

```{r}
data %>% #and then
  select(name, 
         height, 
         gender, 
         species)
```

```{r}
data %>% #and then
  select(1:4)
```

* With `mutate()` you can create new variables or modify existing ones. Let's say we want to create a `height` to `mass` ratio and we might want to express `height` in meters.

```{r}
data %>% #and then
  select(1:3) %>% #and then
  mutate(ratio = height/mass,
         height = height/100)
```

* With `summarise()` you can obtain a succinct summary of one or more variables. 

```{r}
data %>% 
  group_by(sex) %>% 
  summarise(min_height = min(height),
            avg_height = mean(height),
            max_height = max(height)
          )
```

