---
title: "ggplot2 introduction"
author: "Max Kruisbrink"
date: "`r Sys.Date()`"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# define knitr paths
base.dir <- "~/Developer/mkruisbrink.github.io/"
base.url <- "/"
fig.path <- "img/rmd/"

# set knitr parameters
opts_knit$set(base.dir = base.dir, base.url = base.url)
opts_chunk$set(fig.path = fig.path) 
```

## The tidyverse

I wrote [a simple introductory to the tidyverse]({% post_url 2022-08-11-tidyverse-introduction %}) of which `ggplot2` is a core part.
The tidyverse is a package in R that is - in itself - a collection of individual R packages that all work closely together in the same way (they speak thesame kind of language, if you will). Data import & export, manipulation, exploration and visualization becomes simpler and more elegant by using new and additional syntax which is more intuitive to and readable for us humans.

### Set-up

You can install the tidyverse via **`install.packages("tidyverse")`** and you only have to install it once. Lets load the tidyverse and the data we're going to use.

```{r install and load, warning=FALSE}

# installs the package if required
#install.packages("tidyverse")

# loads the tidyverse
library(tidyverse)

```

### Example data

```{r loading data, warning=FALSE}

# using the starwars dataset
data <- starwars

# for the sake of this introduction we will drop all missing values for now
data <- drop_na(data)

```

## ggplot2

```{r}

```

